
title: "Dimentionality Reduction"

author: "Temesgen"

date: "December 3, 2017"

output: html_document

#Chapter 5: Dimentionality Reduction


*Read data from local folder*

```{r}
human <-read.table("D:/GitHub/IODS-project/human.csv", sep = ";", header = TRUE)
str(human)
dim(human)
```

*The data contains information about human development index from most countreis in the world. In this dataset, a total of 8 variables(indicators) and 155 observations are included. Description of the variabels(which are related to health and knowledge, and empowerment) are given below*

*"GNI" = Gross National Income per capita*

*#"Life.Exp" = Life expectancy at birth*

*#"Edu.Exp" = Expected years of schooling*

*#"Mat.Mor" = Maternal mortality ratio*

*#"Ado.Birth" = Adolescent birth rate*

*#"Parli.F" = Percetange of female representatives in parliament*

*#"Edu2.F" = Proportion of females with at least secondary education*

*#"Edu2.M" = Proportion of males with at least secondary education*

*#"Labo2.F" = Proportion of females in the labour force*

*#"Labo2.M" " Proportion of males in the labour force*

*#"Edu2.FM" = Edu2.F/Edu2.M*

*#"Labo.FM" = Labo2.F/Labo2.M*

*Show a graphical overview of the data and summary*

```{r}
library(GGally)
library(dplyr)
library(corrplot)

ggpairs(human)
cor(human)%>% corrplot()
summary(human)
```

*The data distibutions shows the minimum, median, mean, max, 1st and 2nd quantiles. In some countries it shows forexample no representation of female in the parliament(Minmum zero) and big difference in GNI (Max=123124, Min=581). The correlation plot, on the otherhand, shows both negative and postive correlations between variables(e.g. "Edu2.FM" and Mat.Mor shows negative correlation while Mat.Mor and Ado.Birth indicates postitive correlation)*

*Perform principal component analysis (PCA)*

```{r}
library(dplyr)
library(GGally)
library(MASS)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human)

# draw a biplot of the principal component representation and the original variables

biplot(pca_human, choices = 1:2,cex=c(0.8,1),col=c("grey40", "deeppink2"))


```

```{r}
#standardize the variables and repeat the analysis

human_std <- scale(human)


pca_human <- prcomp(human_std)
s <- summary(pca_human)
pca_pr <- round(100*s$importance[2, ], digits = 1)

# create object pc_lab to be used as axis labels
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])

```

*after standrdization the mean value for all variables became zero. And all values to the left and right of the mean have negative and positive respectively. since the pricipal compoents are sensitive to the realtive scaling of variables with large variance(e.g the GNU data(Max=123124, Min=581), it assumes that those with big variance are more important than others. As a result, GNU values are more important than others before standardization. The angle between arrows shows how much they are correlated, the smaller the angle the stronger the correlation is. Forexample, maternal mortality and adolescent birth rate are very close to eaachother, meaning they are correlated. Also, the direction of the arrows shows to which PC-axis they are more important(e.g,Percetange of female representatives in parliament and the proportion of female to male in the labor force are correlated to eachother and are parallel to the PC2  unlike the others). The lenghth of the arrows shows the standard deviations from the mean value*

* the first two principal components capture much of the variation(53.6%) in the human data. As we can see, 6 of the varibles are captured by the first PC1(which makes up most of the variation in the data) followed by PC2( 16.2%). Thus, the data dimensionality is reduced significantly from eight to two unclorrelated principal components*

```{r}
library(FactoMineR)
library(ggplot2)
library(dplyr)
library(tidyr)

data(tea)
keep_columns <- c("Tea", "evening", "price", "tearoom", "where", "sugar")
tea_time <-dplyr::select(tea, one_of(keep_columns))

# look at the summaries and structure of the data
summary(tea_time)
str(tea_time)

# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free")+ geom_bar() +theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

mca <- MCA(tea_time, graph = FALSE)

# summary of the model

summary(mca)
# visualize MCA
plot(mca, invisible=c("ind"), habillage="quali")
plot(mca, invisible=c("var"), habillage="quali")

```

*summaries of the results of the Multiple Correspondence Analysis shows for all categoris the coordinates of the values, the contribution on the axis and the quality of representation measured by square of the cosine( which follows a gaussian distribution) , the correlation ratio between variable(et2). Thus forexample black tea has coordinate 0.477 on the first axis (dim1), and its contribution is 3.159, and the v.test shows that black tea has large positive value(> 1.96) in the first and second dimension while large negative value(< -1.96) in the third dimension*
